Mapping mutational effects along the evolutionary landscape of HIV envelope

### 2018-02-23 paper/manuscript.tex
discussed expression level and limiting to ectodomain and transmembrane domain

Both BG505 and BF520 are from clade A of the major (M) group of HIV-1.
\FIG{tree} shows the phylogenetic relationship among these two Envs and other clade A sequences.
BG505 and BF520 are identical at 721 of the 836 pairwise-alignable protein sites (86.2\% identity).
In our experiments, we analyze only the ectodomain and transmembrane domain of Env (we exclude the signal peptide and cytoplasmic tail).
In these regions of Env, BG505 and BF520 are identical at 549 of the 616 sites (89.1\% identity) that are alignable across clade A Envs (\FIG{tree}-source~data~\ref{figdata:alignment}, \FIG{tree}-source~data~\ref{figdata:mask}).
The divergence between BG505 and BF520 therefore offers ample opportunity to investigate mutational shifts during Env evolution.

@@

We can use the null distribution to identify sites where the shifts between BG505 and BF520 are significantly larger than the noise in our experiments (\FIG{prefsdist}B).
There are 30 such sites at a false discovery rate of 0.1.
\FIG{prefsdist}C shows the amino-acid preferences of these significantly shifted sites for each Env.
For the majority of shifted sites, one Env prefers a specific amino acid whereas the other Env tolerates many amino acids; for instance, see sites 512, 516, 599, 165, 605 and 505 in \FIG{prefsdist}C.
Such broadening and narrowing of a site's mutational tolerance is frequently linked to changes in protein stability, with a more stable protein typically being more mutationally tolerant~\citep{wang2002evolution,bloom2006protein,gong2013stability,kumar2017stability}.
Work with engineered Env protein in the form of ``SOSIP'' trimer~\citep{binley2000recombinant,sanders2002stabilization} has shown that BG505 SOSIP is more thermostable than BF520 SOSIP~\citep{verkerke2016epitope}.
Consistent with this fact, sites with altered mutational tolerance are often (although not always, see sites 165 and 520 in \FIG{prefsdist}C) more mutationally tolerant in BG505.

@@

\subsection{Creation of codon-mutant libraries}
Our codon mutant libraries mutagenized all sites in \textit{env} to all 64 codons, except that the signal peptide and cytoplasmic tail were not mutagenized.
The rationale for excluding these regions is that they are not part of Env's ectodomain, and may be prone to mutations that strongly modulate Env's expression level.

>>>>

Both BG505 and BF520 are from clade A of the major (M) group of HIV-1.
\FIG{tree} shows the phylogenetic relationship among these two Envs and other clade A sequences.
BG505 and BF520 are identical at 721 of the 836 pairwise-alignable protein sites (86.2\% identity).
However, in our experiments we mutagenized only the ectodomain and transmembrane domain of Env, and excluded the signal peptide and cytoplasmic tail.
The reason is that we measure how Env mutations affect viral growth, which is influenced both by the functionality of Env protein molecules and their expression level.
Mutations in the signal peptide and cytoplasmic tail commonly affect Env expression level~\citep{chakrabarti1989cytoplasmic,yuste2004modulation,li1994control}, so we excluded these regions with the goal of reducing the degree to which we simply identified mutations that affected Env expression.
In the ectodomain and transmembrane domains of Env, BG505 and BF520 are identical at 549 of the 616 sites (89.1\% identity) that are alignable across clade A Envs (\FIG{tree}-source~data~\ref{figdata:alignment}, \FIG{tree}-source~data~\ref{figdata:mask}).
The divergence between BG505 and BF520 therefore offers ample opportunity to investigate mutational shifts during Env evolution.

@@

We can use the null distribution to identify sites where the shifts between BG505 and BF520 are significantly larger than the noise in our experiments (\FIG{prefsdist}B).
There are 30 such sites at a false discovery rate of 0.1.
\FIG{prefsdist}C shows the amino-acid preferences of these significantly shifted sites for each Env.
For the majority of shifted sites, one Env prefers a specific amino acid whereas the other Env tolerates many amino acids; for instance, see sites 512, 516, 599, 165, 605 and 505 in \FIG{prefsdist}C.
Such broadening and narrowing of a site's mutational tolerance is frequently linked to changes in protein stability, with a more stable protein typically being more mutationally tolerant~\citep{wang2002evolution,bloom2006protein,gong2013stability,kumar2017stability}.
Work with engineered Env protein in the form of ``SOSIP'' trimer~\citep{binley2000recombinant,sanders2002stabilization} has shown that BG505 SOSIP is more thermostable than BF520 SOSIP~\citep{verkerke2016epitope}.
Consistent with this fact, sites with altered mutational tolerance are often (although not always, see sites 165 and 520 in \FIG{prefsdist}C) more mutationally tolerant in BG505.
Differences in Env's expression level might also contribute to a general broadening or narrowing of tolerance to subsequent mutations.
Because our experiments select for viral growth---which is affected by both Env function and expression---it is possible that some of the shifts are due to epistatic mutational effects on expression rather than function.

@@

\subsection{Creation of codon-mutant libraries}
Our codon mutant libraries mutagenized all sites in \textit{env} to all 64 codons, except that the signal peptide and cytoplasmic tail were not mutagenized.
The rationale for excluding these regions is that they are not part of Env's ectodomain, and are prone to mutations that strongly modulate Env's expression level~\citep{chakrabarti1989cytoplasmic,yuste2004modulation,li1994control}.
